Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-05-19 main.Rmd
subheader changed to computation only

## Computation through calendar algebra

@@

## Computation through cyclic calendar algebra

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple** and (2) **multiple-to-single** which entails the representation of _multiple-order-up_ cyclic granularities from _single-order-up_ cyclic granularities respectively.

@@

\noindent All cyclic granularities can be expressed in terms of the index set, and hence, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Let $N_C$ be the total number of contextual circular, quasi-circular and aperiodic granularities. Any attempt to encode all or many of these cyclic granularities at the same time to develop insights on periodicity might fail or otherwise become too numerous for comprehensive human consumption. Instead, this big problem is broken down by focusing on pair of cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

```{r data-structure, fig.cap = "The data structure for exploring periodicities in data by including cyclic granularities in the tsibble structure with index, key and measured variables.", fig.pos="htb"}
include_graphics("Figs/data-structure-v2.png")
```

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. Here, $A_i$'s or $B_j$'s are the levels/categories corresponding to $C_1$ and $C_2$ respectively. Let $S_{ij}$ be a subset of the index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_1$ be day-of-month with 31 levels and $C_2$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to possible combinations of $C_1$ and $C_2$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month. Hence the pair (day-of-month, week-of-month) is a clash.

<!--  In fact, most of these 155 sets will be empty, making the combination of $C_1$ and $C_2$ in a graph unhelpful. -->

Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Let us consider $C_1$ be day-of-week with 7 levels and $C_2$ be WorkingDay/NonWorkingDay with 2 levels. While potentially all of these 14 sets $S_{ij}$ can be non-empty (it is possible to have a public holiday on any day-of-week), in practice many of these will probably have very few observations. For example, there are few (if any) public holidays on Wednesdays or Thursdays in any given year in Melbourne, Australia.

<!-- However, this is different from the first example because these are not structurally empty. Also, the graphic may still be useful if we want to look at whether the interaction of $C_1$ and $C_2$ affects the distribution we are plotting. -->

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularities does.

An example when there will be no empty combinations could be where $C_1$ and $C_2$ maps index set to day-of-week and month-of-year respectively. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.

@@

Even after screening harmonies from the list of all possible combinations of cyclic time granularities that could be plotted together, the number of combinations quickly becomes too large and thus, it is beneficial to rank these harmonies. Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth.

<!-- Calculating distances and lit review -->

<!-- Normalizing -->

 Distances between these probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leibler divergence measure.

1. Suppose $C_1$ and $C_2$ are two cyclic granularities such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ is the measured variable. Hence for each combination ($A_i$, $B_j$), we have the time series variable $v_{ij} \subseteq v$, $\forall i = {1, 2, \dots, l}$ and $\forall j = {1, 2, \dots, m}$

2. There will be $m$ time series variable corresponding to each level $A_i$ of $C_1$. We compute the pairwise differences in distributions of these $m$ time series variables. Thus there will be  $\binom{m}{2}$ pairwise distances corresponding to each $A_i$.

3. Pairwise distances are computed between different quantiles of the time series variables using Jensen-Shanon distance. Suppose $q$ is the sample quantile vector computed for probability vector $p$. Then Jensen-Shanon distance between two sample quantiles $q_j$ and $q_k$ will be given by $$d_{jk} = [D(q_j, r) + D(q_k, r)]/2 \quad where \quad r = (q_j + q_k)/2$$ where,
 $$D(q_j,q_k) = \int^{\infty}_{-\infty}q_j(x)log\frac{q_j(x)}{q_k(x)}\,dx$$ is the Kullback-Leibler divergence between $q_j$ and $q_k$.

@@

6. Steps 1 to 5 are repeated for every harmony pair and harmony pairs rearranged from highest to lowest MMPD. Harmony pair with higher MMPD would be more interesting as it would mean that the harmony pair exhibits more variations in distributions on an average between different levels of $C_2$.

@@

<!-- The three key components of every visualization is: data, aesthetic mappings between variables in the data and visual attributes, and at least one layer which describes how to render each observation. Layers are usually created with a geom function. Considering the data structure <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$, visualizations can be defined by: -->
<!-- - Data: <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ -->
<!-- - Aesthetic mapping: $A_i$ mapped to $x$ position and $v(s)$ to $y$ position. -->
<!-- - Layer: any geom that summarizes distribution -->

@@

<!-- We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate cyclic temporal granularities with the data structure <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$. The framework involves faceting approach with one cyclic granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

@@

We will discuss the effect of the levels of cyclic granularities on the choice of plots <!--Recall that for cyclic granularities $C_1$ and $C_2$ such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. $A_i$'s and $B_j$'s are referred to as the levels of $C_1$ and $C_2$ for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ respectively.The levels have an impact on the--> since space and resolution might become a problem if the number of levels are too high. A potential approach could be to categorize the levels as very high/high/medium/low for each cyclic granularity and define some criteria based on usual cognitive power, display size available and the aesthetic mappings. Default values for these categorizations could be chosen based on levels of common temporal granularities like days of the month, days of the fortnight or days of the week.<!--For example, any levels above 31 can be considered as very high, any levels between 14 to 31 can be taken as high and that between 7 to 14 can be taken as medium and below 7 as low.>

>>>>

## Computation

@@

## Computation

In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities. This is similar to calendar algebra in @Ning2002-tf for linear granularities but caters to computation of cyclic granularities, we shall refer to it as through cyclic calendar algebra operations. The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple** and (2) **multiple-to-single** which entails the representation of _multiple-order-up_ cyclic granularities from _single-order-up_ cyclic granularities respectively.

@@

\noindent All cyclic granularities can be expressed in terms of the index set, and hence, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ periodic linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Let $N_C$ be the total number of contextual circular, quasi-circular and aperiodic granularities that could originate from $N_L$ periodic and aperiodic linear granularities. Any attempt to encode all or many of these cyclic granularities at the same time to develop insights on periodicity might fail or otherwise become too numerous for comprehensive human consumption. Instead, this big problem is broken down by focusing on pair of cyclic granularities ($C_i$, $C_j$) at a time and data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable $v$.

@@

```{r data-structure, fig.cap = "The data structure for exploring periodicities in data by including cyclic granularities in the tsibble structure with index, key and measured variables.", fig.pos="htb"}
include_graphics("Figs/data-struc-rev.png")
```

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_i$ and $C_j$, such that $C_i$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_I$\}, and $C_j$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_J$\}. Here, $A_k$'s or $B_l$'s are the levels/categories corresponding to $C_i$ and $C_j$ respectively. Let $S_{kl}$ be a subset of the index set such that for all $s \in S_{kl}$, $C_i(s) = A_k$ and $C_j(s) = B_l$. Data subsets for each combination of levels ($A_k$, $B_l$) like <$A_k$, $B_l$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy. These are called "structurally" empty combinations. Let us take a specific example, where $C_i$ be day-of-month with 31 levels and $C_j$ be week-of-month with 5 levels. There will be $31\times 5=155$ sets $S_{kl}$ corresponding to possible combinations of $C_i$ and $C_j$. Many of these like $S_{1,5}$, $S_{21,2}$ are empty. This is also intuitive since the first day of the month can never correspond to fifth week of the month. Hence the pair (day-of-month, week-of-month) is a clash.

<!--  In fact, most of these 155 sets will be empty, making the combination of $C_i$ and $C_j$ in a graph unhelpful. -->

Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Let us consider $C_i$ be day-of-week with 7 levels and $C_j$ be WorkingDay/NonWorkingDay with 2 levels. While potentially all of these 14 sets $S_{kl}$ can be non-empty (it is possible to have a public holiday on any day-of-week), in practice many of these will probably have very few observations. For example, there are few (if any) public holidays on Wednesdays or Thursdays in any given year in Melbourne, Australia.

<!-- However, this is different from the first example because these are not structurally empty. Also, the graphic may still be useful if we want to look at whether the interaction of $C_i$ and $C_j$ affects the distribution we are plotting. -->

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_i$ be Business-days, which are days from Monday to Friday except holidays and $C_j$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularities does.

An example when there will be no empty combinations could be where $C_i$ and $C_j$ maps index set to day-of-week and month-of-year respectively. Here $C_i$ can have 7 levels while $C_j$ can have 12 levels. So there are $12\times7=84$ sets $S_{kl}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.

@@

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granulairites. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $N_C \choose 2$. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left could be large and overwhelming for human consumption. Hence, there is a need to rank these harmonies basis how well they capture the variation in the measured variable. The harmony pair $(C_i, C_j)$ would be most interesting if the average variation in the measured variable between different levels of $C_i$ and $C_j$ is maximum.<!-- Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. -->

<!-- Calculating distances between distributions -->

\noindent One of the potential ways to evaluate this variation is by computing the pairwise distances between the distributions of the measured variable across different levels of $C_j$ and integrating it for all levels of $C_i$. Distances or similarity measures have been extensively studied and applied to fields like clustering, classification and anomaly detection. We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence which is based on Kullback-Leibler divergence. In our work, probability distributions are represented through sample quantiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. Suppose $q$ is a sample quantile vector of the measured variable computed for probability vector $p$. Then Jensen-Shanon distance between two sample quantiles $q_j$ and $q_k$ will be given by $$d_{jk} = [D(q_j, r) + D(q_k, r)]/2 \quad where \quad r = (q_j + q_k)/2$$ where,
 $$D(q_j,q_k) = \int^{\infty}_{-\infty}q_j(x)log\frac{q_j(x)}{q_k(x)}\,dx$$ is the Kullback-Leibler divergence between $q_j$ and $q_k$.

<!-- Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. -->

<!-- # distribution of distances -->
<!-- # Normalizing and lit review -->

\noindent Since we are looking to rank the harmony pairs based on if it captures the maximum variation across levels of $C_i$ and $C_j$, we need to make sure that this maximum value is independent of the number of levels of the two cyclic granularities. A major result in the field of EVT is the probability distribution of the maximum can only be one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process. But the normalising constants vary depending on the underlying distribution of the measured variables. Hence to normalize appropriately, it is important to assume a distribution of distances in our case.

<!-- For ranking the harmony pairs, these distances need to be normalised by the number of levels of . -->

<!-- Distances between these probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leibler divergence measure. -->

1. Suppose $C_1$ and $C_2$ are two cyclic granularities such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ is the measured variable. Hence for each combination ($A_k$, $B_j$), we have the time series variable $v_{ij} \subseteq v$, $\forall i = {1, 2, \dots, l}$ and $\forall j = {1, 2, \dots, m}$

2. There will be $m$ time series variable corresponding to each level $A_k$ of $C_1$. We compute the pairwise differences in distributions of these $m$ time series variables. Thus there will be  $\binom{m}{2}$ pairwise distances corresponding to each $A_k$.

3. Pairwise distances are computed between different quantiles of the time series variables using Jensen-Shanon distance.

<!-- Suppose $q$ is the sample quantile vector computed for probability vector $p$. Then Jensen-Shanon distance between two sample quantiles $q_j$ and $q_k$ will be given by $$d_{jk} = [D(q_j, r) + D(q_k, r)]/2 \quad where \quad r = (q_j + q_k)/2$$ where, -->
<!--  $$D(q_j,q_k) = \int^{\infty}_{-\infty}q_j(x)log\frac{q_j(x)}{q_k(x)}\,dx$$ is the Kullback-Leibler divergence between $q_j$ and $q_k$. -->

@@

6. Steps 1 to 5 are repeated for every harmony pair and harmony pairs rearranged from highest to lowest MMPD.

<!-- with higher MMPD would be more interesting as it would mean that the harmony pair exhibits more variations in distributions on an average between different levels of $C_2$. -->

@@

<!-- The three key components of every visualization is: data, aesthetic mappings between variables in the data and visual attributes, and at least one layer which describes how to render each observation. Layers are usually created with a geom function. Considering the data structure <$A_k$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$, visualizations can be defined by: -->
<!-- - Data: <$A_k$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ -->
<!-- - Aesthetic mapping: $A_k$ mapped to $x$ position and $v(s)$ to $y$ position. -->
<!-- - Layer: any geom that summarizes distribution -->

@@

<!-- We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate cyclic temporal granularities with the data structure <$A_k$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$. The framework involves faceting approach with one cyclic granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

@@

We will discuss the effect of the levels of cyclic granularities on the choice of plots <!--Recall that for cyclic granularities $C_1$ and $C_2$ such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. $A_k$'s and $B_j$'s are referred to as the levels of $C_1$ and $C_2$ for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ respectively.The levels have an impact on the--> since space and resolution might become a problem if the number of levels are too high. A potential approach could be to categorize the levels as very high/high/medium/low for each cyclic granularity and define some criteria based on usual cognitive power, display size available and the aesthetic mappings. Default values for these categorizations could be chosen based on levels of common temporal granularities like days of the month, days of the fortnight or days of the week.<!--For example, any levels above 31 can be considered as very high, any levels between 14 to 31 can be taken as high and that between 7 to 14 can be taken as medium and below 7 as low.>
